---
title: "UICC Stage Rules"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
pkgload::load_all(".")
library(dplyr)
```

## Overview

Internal auxiliary files are used to determine cancer stage.

## Details

The package stores three rule sets derived from UICC guidelines for the TNM staging system. The rules stored in these files are differentiated by:

- **Cancer site**: bladder, breast, colorectal, lung, oesophageal, prostate, skin
- **Classification type**: clinical, pathological, base
- **Classification edition**: 7th, 8th, 9th, unspecified

The internal data is divided in two categories *concepts* and *mapping* rules. 

## Concepts (tnm_concepts)

Contains information about the individual TNM components and their concept IDs, differentiated by type and edition.

There are **393 total concepts** for **44 unique components**:

```{r, echo=FALSE}
tnm_components <- c(
  "TX", "T0", "Tis", "Ta", "T1", "T1a", "T1b", "T1c", "T1mi",
  "T2", "T2a", "T2b", "T2c", "T3", "T3a", "T3b", "T4", "T4a",
  "T4b", "T4c", "T4d", "NX", "N0", "N1", "N1a", "N1b", "N1c",
  "N1mi", "N2", "N2a", "N2b", "N2c", "N3", "N3a", "N3b", "N3c",
  "M0", "M1", "M1a", "M1b", "M1c", "M1c1", "M1c2", "M1d"
)

paste(tnm_components, collapse = ", ")
```

These components have different versions according to edition and type.

### By edition

| Edition | Number of concepts |
|---|---:|
| 7th edition | 131 |
| 8th edition | 131 |
| Unspecified edition | 131 |

### By type

| Type | Number of concepts |
|---|---:|
| Base | 132 |
| Clinical | 132 |
| Pathological | 129 |

> The `M0` component is not valid in the pathological setting for any edition.

## Mapping (tnm_mapping)

Contains the rules used to determine cancer stage based on a combination of individual TNM components.

Rules are differentiated by:

| Rule dimension | Values |
|---|---|
| Cancer site | Bladder, breast, colorectal, lung, oesophageal, prostate, skin |
| Classification type | Clinical, pathological, base |
| Classification edition | 7th, 8th, 9th |

Each rule refers to a specific source page of the UICC guidelines.

The currently available rules support the following concept categories.

```{r, echo=FALSE}
stage_mapping <- readStagesRDS("mapping")

available_rules <- stage_mapping |>
  count(
    site,
    edition,
    stage_grouping_scope,
    name = "number_of_rules"
  ) |>
  filter(number_of_rules > 0) |>
  arrange(site, edition, stage_grouping_scope)

available_rules |>
  rename(
    `Cancer site` = site,
    Edition = edition,
    `Classification type` = stage_grouping_scope,
    `Number of rules` = number_of_rules
  ) |>
  knitr::kable()
```

Browse the full rule sets below.

```{r, echo=FALSE, results='asis'}
format_label <- function(x) {
  paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}

for (edition_value in unique(available_rules$edition)) {
  cat(sprintf(
    "<details>\n<summary>%s edition</summary>\n\n",
    htmltools::htmlEscape(edition_value)
  ))

  edition_rules <- available_rules |>
    filter(edition == edition_value)

  for (site_value in unique(edition_rules$site)) {
    cat(sprintf(
      "<details style=\"margin-left: 1rem;\">\n<summary>%s</summary>\n\n",
      htmltools::htmlEscape(format_label(site_value))
    ))

    site_rules <- edition_rules |>
      filter(site == site_value)

    for (scope_value in unique(site_rules$stage_grouping_scope)) {
      rules <- stage_mapping |>
        filter(
          edition == edition_value,
          site == site_value,
          stage_grouping_scope == scope_value
        ) |>
        select(T, N, M, uicc_stage) |>
        arrange(uicc_stage, T, N, M)

      cat(sprintf(
        "<details style=\"margin-left: 1rem;\">\n<summary>%s (%s rules)</summary>\n\n",
        htmltools::htmlEscape(format_label(scope_value)),
        nrow(rules)
      ))

      cat(knitr::kable(rules, format = "html"))
      cat("\n</details>\n\n")
    }

    cat("</details>\n\n")
  }

  cat("</details>\n\n")
}
```

## References

Sobin, L. H., Gospodarowicz, M. K., & Wittekind, C. (Eds.). (2009). *TNM classification of malignant tumours* (7th ed.). Wiley-Blackwell.

Brierley, J. D., Gospodarowicz, M. K., & Wittekind, C. (Eds.). (2016). *TNM classification of malignant tumours* (8th ed.). Wiley-Blackwell.

Brierley, J. D., Giuliani, M., O'Sullivan, B., Rous, B., & Van Eycken, E. (Eds.). (2025). *TNM classification of malignant tumours* (9th ed.). Wiley.